---
title: "A Typical pathfindR Workflow"
author: "Ege Ulgen"
date: "2026-07-07"
slug: pathfindr-typical-workflow
categories:
  - R
  - bioinformatics
tags:
  - tutorial
  - enrichment-analysis
  - pathways
  - active-subnetworks
output:
  blogdown::html_page:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 8,
  fig.height = 6
)
```

[pathfindR](https://github.com/egeulgen/pathfindR/) is an R package for
pathway/gene-set enrichment analysis that does something most enrichment tools
don't: instead of testing your gene list directly, it first maps your genes onto
a protein–protein interaction network (PIN), finds **active subnetworks**
(connected groups of interacting genes that are enriched for significantly
altered genes), and then runs enrichment on those subnetworks. The payoff is that
genes which don't pass your significance cutoff on their own (but sit in the
middle of a cluster of genes that do) still get a chance to contribute.

This post walks through a typical workflow end to end: run the analysis, cluster
the results, visualize, and score terms per sample.

## What changed in 3.0.0: no more Java

If you used older versions, you'll remember the asterisk next to "easy install":
the active subnetwork search ran on a bundled Java `.jar`, so you needed a JRE on
your `PATH` and a fair amount of "is Java configured correctly?" troubleshooting.

As of **pathfindR 3.0.0**, that's gone. The greedy (GR), simulated annealing
(SA), and genetic (GA) search algorithms have been re-implemented in R and C++
(via Rcpp), and the Java dependency was removed entirely. For GR and SA, results
are numerically identical to the legacy JAR given the same seed and inputs so
this is a drop-in upgrade, not a behaviour change. Installation is now just:

```{r install, eval=FALSE}
install.packages("pathfindR")
```

That's the whole story. No `JAVA_HOME`, no PATH surgery.

## The input

The input is a simple data frame with three columns: a gene symbol, an
(optional) change value such as log-fold-change, and a p-value. pathfindR ships
with an example derived from a rheumatoid arthritis vs. healthy
differential-expression dataset (GEO `GSE15573`):

```{r input}
library(pathfindR)

head(example_pathfindR_input)
```

## Running the analysis

The wrapper `run_pathfindR()` takes you from input to enriched terms in one call:
it processes the input, maps it onto the PIN, searches for active subnetworks,
and runs enrichment over a chosen number of iterations.

```{r run, eval=FALSE}
output_df <- run_pathfindR(example_pathfindR_input)
```

A few arguments worth knowing (all optional):

```{r run-args, eval=FALSE}
# pick the gene sets (default = "KEGG")
output_df <- run_pathfindR(example_pathfindR_input, gene_sets = "GO-MF")

# pick the PIN for the subnetwork search (default = "Biogrid")
output_df <- run_pathfindR(example_pathfindR_input, pin_name_path = "STRING")

# write the HTML report somewhere specific
output_df <- run_pathfindR(example_pathfindR_input, output_dir = "RA_results")
```

Available PINs include `"Biogrid"`, `"STRING"`, `"GeneMania"`, `"IntAct"`,
`"KEGG"` and `"mmu_STRING"`; available gene sets include `"KEGG"`, `"Reactome"`,
`"BioCarta"`, the GO families (`"GO-All"`, `"GO-BP"`, `"GO-CC"`, `"GO-MF"`) and
`"mmu_KEGG"`. You can also supply a custom PIN or custom gene sets.

The result is a data frame, one row per enriched term, with the term ID and
description, enrichment p-values (lowest/highest across iterations), fold
enrichment, and the affected up-/down-regulated input genes. The package also
bundles a pre-computed copy of this output so we can keep going without waiting
on the search:

```{r output}
head(example_pathfindR_output, 4)
```

A quick way to eyeball the top hits is the enrichment chart, a bubble plot of the
most significant terms:

```{r enrichment-chart}
enrichment_chart(example_pathfindR_output, top_terms = 15)
```

## Clustering the enriched terms

Enrichment output is often redundant, several terms describe overlapping
biology. `cluster_enriched_terms()` computes pairwise kappa statistics between
terms, clusters them (hierarchical by default), picks the cluster count that
maximizes the average silhouette width, and tags one **representative** term per
cluster:

```{r cluster}
clustered_df <- cluster_enriched_terms(example_pathfindR_output)

# the representative term from each cluster
clustered_df[clustered_df$Status == "Representative", c("Term_Description", "Cluster")]

enrichment_chart(example_pathfindR_output, plot_by_cluster = TRUE)
```

Filtering to representatives gives you a compact, non-redundant shortlist to
report. There's also a `method = "fuzzy"` option if you'd rather let terms belong
to more than one cluster.

## Visualizing results

Beyond the enrichment chart, two visuals are especially handy.

A **term–gene heatmap** shows which input genes drive which terms. If you pass
the original input, tiles are colored by change value:

```{r heatmap}
term_gene_heatmap(example_pathfindR_output, example_pathfindR_input, num_terms = 10)
```

A **term–gene graph** shows the same relationships as a network, making shared
genes between terms obvious. As of 3.x this is split into a builder and a
plotter:

```{r term-gene-graph}
g <- create_term_gene_graph(example_pathfindR_output, num_terms = 10)
create_term_gene_plot(g)
```

## Scoring terms per sample

If you have a per-sample expression matrix, `score_terms()` computes an
agglomerated z-score for each enriched term in each sample — useful for seeing
whether a pathway is collectively activated or repressed in particular samples or
groups:

```{r score}
case_sample_ids <- c(
  "GSM389703", "GSM389704", "GSM389705", "GSM389706", "GSM389707", "GSM389708"
)
score_matrix <- score_terms(
  enrichment_table = clustered_df[clustered_df$Status == "Representative", ],
  exp_mat          = example_experiment_matrix,   # genes x samples
  cases            = case_sample_ids,             # optional
  plot_hmap        = TRUE
)
```

## Putting it together

The whole everyday workflow is really just:

```{r full, eval=FALSE}
library(pathfindR)

output_df    <- run_pathfindR(example_pathfindR_input)          # enrichment
clustered_df <- cluster_enriched_terms(output_df)               # de-duplicate
enrichment_chart(clustered_df, plot_by_cluster=TRUE)            # visualize
term_gene_heatmap(output_df, example_pathfindR_input)           # who drives what
```

Four functions, and no Java to install. If you're comparing two
conditions, take a look at `combine_pathfindR_results()` too.

---

*pathfindR is described in Ulgen E, Ozisik O, Sezerman OU. 2019.
pathfindR: An R Package for Comprehensive Identification of Enriched Pathways in
Omics Data Through Active Subnetworks. Front. Genet.
<https://doi.org/10.3389/fgene.2019.00858>. Docs:
<https://egeulgen.github.io/pathfindR/>.*
